---
title: "How does high tempuratures affect mental health in Baltimore City?"
format: gfm
editor: visual
---

Step 1. Install necessary packages.

```{r}
#| eval: false
#| warning: false
#| message: false
install.packages("tidyverse")
install.packages("kableExtra")

install.packages("terra")
```

Step 2. Declare that you will use these packages in this session.

```{r}
#| warning: false
#| message: false
library("tidyverse")
library("kableExtra")
```

Step 3. Upload the dataframe that you have created in Spring 2024 into the repository.

Step 4. Open the dataframe into the RStudio Environment.

```{r}
df<-read.csv("panel.csv")
df2<-df %>%
  mutate(call_bin = ifelse(callscount>0, 1, 0)) %>%
  mutate(date = as.Date(date)) %>%
  mutate(month = month(date)) %>%
  mutate(dow = weekdays(date))
```

Step 5. Use the **head** and **kable** function showcase the first 10 rows of the dataframe to the reader.

```{r}
kable(head(df2))
```

## Introduction & Literature Review:

Understanding how temperature affects mental health is crucial for society. With rising global temperatures, we will have to be prepared for any impacts that may cause. The years from 2013 to 2023 have been the warmest to date, with 2023 having the highest surface temperature ever recorded (NASA, n.d). High temperatures were studied in Baltimore city through in depth interviews where it was found that at higher heats people were more prone to mental health problems such as depression and anxiety. Especially disadvantaged individuals such as people who are low-income, have disabilities, are homeless, pregnant, elderly people and children (Diallo et al., 2024). If we were to discover more impacts that temperature has on individuals, it could guide urban planners to plan future cities so that they would include public spaces throughout the city. Additionally, by understanding the psychological effects of temperature, governments and healthcare providers can better prepare interventions, such as improving access to mental health services during extreme temperatures. This research could also lead to targeted policies that aim to mitigate the mental health impacts of climate change, ultimately improving societal well-being.

This research project aims to provide the most concentrated area example of the impacts of temperature on mental health. While existing studies have found links between mental health worsening and high temperatures, most have done so for the whole country of the US by county, which makes it more difficult to definitively control for other factors (Burke et al., 2018; Srivastava & Mullins., 2024). The data used by Burke et al. (2018) is based on suicide rates and is from 1968–2004 in the US and 1990-2010 in Mexico by county and analyzed monthly. They found a 0.7% suicide rate increase in the US per 1.8 degrees Fahrenheit increase in monthly temperature. 

Srivastava & Mullins (2024) compared the daily frequencies of crisis line conversations from each county with local average daily temperatures. They found an 8% increase in crisis line conversations on days exceeding 86 degrees Fahrenheit compared to days between 64.4 to 69.8 degrees Fahrenheit. 

Besides existing studies on the link between temperatures and mental health, there are many studies that looked at the impact of temperature on crime rates. Heilmann et al. (2021) discovered that crime rates in Los Angeles police districts increase 1.72% when daily temperature exceeds 75 and 1.90% when daily temperature exceeds 75 and 90 degrees Fahrenheit. 

Our study will differ from existing studies on temperature and mental health on several key points. We will be analyzing a more controlled location and group, focussing on Baltimore City, allowing us to control for factors specific to the police district. We will be using some of the same strategies as other studies (Burke et al., 2018; Srivastava & Mullins., 2024; Heilmann et al., 2021) such as controlling for factors of precipitation, air quality, and holidays. As well as adding an extra factor of our own, greenspace. Our data will be more modern mental health crisis call data from 2021-2023 and analyzed daily to account for factors such as day of week.

**Bibliography:** 

Burke, M., González, F., Baylis, P., Heft-Neal, S., Baysan, C., Basu, S., & Hsiang, S. (2018). Higher temperatures increase suicide rates in the United States and Mexico. Nature Climate Change, 8(8), 723–729. <https://doi.org/10.1038/s41558-018-0222-x> 

Change, N. G. C. (n.d.). Global Surface Temperature \| NASA Global Climate Change. Climate Change: Vital Signs of the Planet. Retrieved October 22, 2024, from <https://climate.nasa.gov/vital-signs/global-temperature/?intent=121> 

Diallo, I., He, L., Koehler, K., Spira, A. P., Kale, R., Ou, J., Smith, G., Linton, S. L., & Augustinavicius, J. (2024). Community perspectives on heat and health in Baltimore City. Urban Climate, 54, 101841. <https://doi.org/10.1016/j.uclim.2024.101841> 

Heilmann, K., Kahn, M. E., & Tang, C. K. (2021). The urban crime and heat gradient in high and low poverty areas. Journal of Public Economics, 197, 104408. <https://doi.org/10.1016/j.jpubeco.2021.104408>

Srivastava, S., & Mullins, J. T. (2024). Temperature, Mental Health, and Individual Crises: Evidence from Crisis Text Line. American Journal of Health Economics. <https://doi.org/10.1086/730332> 

## Question 1: What is the frequency of this data frame?

Answer: Daily- day and night

## Question 2: What is the cross-sectional (geographical) unit of this data frame?

Answer: Police district

Step 6. Use the **names** function to display all the variables (column) in the dataframe.

```{r}
names(df2)

```

## Question 3: Which column represents the treatment variable of interest?

Answer: temp_F

## Question 4: Which column represents the outcome variable of interest?

Answer: call_bin

Step 7: Create a boxplot to visualize the distribution of the outcome variable under treatment and no treatment.

```{r}

# histogram or scatterplots
# total number of calls
# x -> temp_F, y -> callscount (avg for scatterplot)
# groupBy policeDistrict

#----------------------------------------------------------------------------------------------
df2 <- df2 %>%
  mutate(tempCatagories = ifelse(temp_F < 10, 0,
                          ifelse(temp_F >= 10 & temp_F < 20, 10,
                          ifelse(temp_F >= 20 & temp_F < 30, 20,
                          ifelse(temp_F >= 30 & temp_F < 40, 30,
                          ifelse(temp_F >= 40 & temp_F < 50, 40,
                          ifelse(temp_F >= 50 & temp_F < 60, 50,
                          ifelse(temp_F >= 60 & temp_F < 65, 60,
                          ifelse(temp_F >= 65 & temp_F < 70, 65,
                          ifelse(temp_F >= 70 & temp_F < 75, 70,
                          ifelse(temp_F >= 75 & temp_F < 80, 75,
                          ifelse(temp_F >= 80 & temp_F < 85, 80,
                          ifelse(temp_F >= 85 & temp_F < 90, 85,
                          ifelse(temp_F >= 90 & temp_F < 95, 90,
                          ifelse(temp_F >= 95 & temp_F < 100, 95, 
                          100)))))))))))))))

# if 0 -> summer (4 - 9)
# if 1 -> winter (11 - 2)
# if 2 -> march and october
df2 <- df2 %>%
  mutate(seasons = ifelse(month >= 4 & month <= 9, 0,
                      ifelse(month <= 2 | month >= 11, 1,
                             2)))
         
#df2_summer <- df2 %>%
         #filter(month>=4 & month<=9)
         
# for winter
winter_df <- df2 %>%
  filter(seasons == 1) %>%
  group_by(policeDistrict, tempCatagories) %>%
  summarise(avg_calls = mean(callscount, na.rm = TRUE))

# for summer
summer_df <- df2 %>%
  filter(seasons == 0) %>%
  group_by(policeDistrict, tempCatagories) %>%
  summarise(avg_calls = mean(callscount, na.rm = TRUE))

# Scatter plot -winter
ggplot(winter_df, aes(x = tempCatagories, y = avg_calls, color = policeDistrict)) + 
  geom_point(size = 2) + 
  geom_smooth(se = FALSE) +
  labs(title = "Average Number of Calls vs Temperature (F) by Police District (Winter 11-2)",
       x = "Temperature Range (F)", 
       y = "Average Number of Calls")

# Scatter plot -summer
ggplot(summer_df, aes(x = tempCatagories, y = avg_calls, color = policeDistrict)) + 
  geom_point(size = 2) + 
  geom_smooth(se = FALSE) +
  labs(title = "Average Number of Calls vs Temperature (F) by Police District (Summer 4-9)",
       x = "Temperature Range (F)", 
       y = "Average Number of Calls")

#----------------------------------------------------------------------------------------------

# for winter
df_winter <- df2 %>%
  filter(winter_bin == 1) %>%
  group_by(policeDistrict, tempCatagories) %>%
  summarize(total_callscount = sum(callscount))

# for summer
df_summer <- df2 %>%
  filter(winter_bin == 0) %>%
  group_by(policeDistrict, tempCatagories) %>%
  summarize(total_callscount = sum(callscount))

# Grouped bar graph -winter
ggplot(df_winter, aes(x = tempCatagories, y = total_callscount)) + 
    geom_bar(stat="identity") +
    facet_wrap((~ as.character(policeDistrict))) +
    labs(title = "Total Number of Calls per Police District (Winter 11-2)",
       x = "Temperature Range (F)", 
       y = "Total Number of Calls")

# Grouped bar graph -summer
ggplot(df_summer, aes(x = tempCatagories, y = total_callscount)) + 
    geom_bar(stat="identity") +
    facet_wrap((~ as.character(policeDistrict))) +
    labs(title = "Total Number of Calls per Police District (Summer 4-9)",
       x = "Temperature Range (F)", 
       y = "Total Number of Calls")
```

\
\

![](data:image/png;base64,iVBORw0KGgoAAAANSUhEUgAAACgAAAAaCAYAAADFTB7LAAAAcElEQVR4Xu3OwQmAQAxE0bClWYCW5N06tM6V2YPg5CjoF/JhLoHAi6iqn9eOefUbqrYvHY0cQDLyAlKRNyARmYA0ZMLRkAlGQyaU72tkAtlim7r/vJqDUDjlKBROOQyFU2icQuMUGqfQuBEaV1XPOwEx96nYACK8+wAAAABJRU5ErkJggg== "Run Current Chunk")

Step 8: Fit a regression model $y=\beta_0 + \beta_1 x + \epsilon$ where $y$ is the outcome variable and $x$ is the treatment variable. Use the **summary** function to display the results.

```{r}
model1<-lm(call_bin ~ temp_F, data=df2)

summary(model1)
```

## Question 7: What is the equation that describes the linear regression above? Please include an explanation of the variables and subscripts.

Answer:

$$
callscount_{pdt} = \beta_0 + \beta_1 temp_{pdt} + \gamma_p + \eta_d + \theta_{dayofweek} + \omega_{month} + \pi_year+\epsilon_{pdt}
$$

Where $callscount$ represents the outcome variable that shows whether there a call from police district \$p\$, on day \$d\$, at time of day \$t\$.

Where $temp$ determines the temperature in Fahrenheit.

Data Description:

-   Outcome variable: whether there was a call or not per day.

-   Treatment variable: temperature

-   Frequency and geographical Unit: daily day and night and police district

-   Treated Group: \# of call on a hot weekday in a specific month in a police district

-   Untreated Group:# of call on a cool weekday in a specific month in a police district

## Question 8: What fixed effects can be included in the regression? What does each fixed effects control for? Please include a new equation that incorporates the fixed effects.

Answer:

Police district: controls for better or worse conditions in the city

Daytime: controls the differences that might occur based on the time of day

Year: controls any differences that might have occurred one year to another

## Question 9: What is the impact of the treatment effect once fixed effects are included?

Answer:

```{r}
#| warning: false
#| message: false
#install.packages("lfe")
library("lfe")

model2<-felm(call_bin ~ temp_F + daytime + temp_F:daytime| 
               policeDistrict + year + month + dow, data=df2)

summary(model2)
```

```{r}
#| warning: false
#| message: false

# for the temperature range
df3<-df2 %>%
  mutate(temp_over_100=ifelse(temp_F>100, 1, 0)) %>%
  mutate(temp_95_100=ifelse(temp_F> 95 & temp_F<=100, 1, 0)) %>%
  mutate(temp_90_95=ifelse(temp_F> 90 & temp_F<=95, 1, 0)) %>%
  mutate(temp_85_90=ifelse(temp_F> 85 & temp_F<=90, 1, 0)) %>%
  mutate(temp_80_85=ifelse(temp_F> 80 & temp_F<=85, 1, 0)) %>%
  mutate(temp_75_80=ifelse(temp_F> 75 & temp_F<=80, 1, 0)) %>%
  mutate(temp_70_75=ifelse(temp_F> 70 & temp_F<=75, 1, 0)) %>%
  mutate(temp_65_70=ifelse(temp_F> 65 & temp_F<=70, 1, 0)) %>%
  mutate(temp_60_65=ifelse(temp_F> 60 & temp_F<=65, 1, 0)) %>%
  mutate(temp_under_60=ifelse(temp_F<= 60, 1, 0))

#-------#

#2021 - 2022 (from the website: https://patch.com/maryland/baltimore/back-school-baltimore-city #-2021-22-school-calendar) (Not full but )
#2022 - 2023
#2023 - 2024 (upto Dec 31)

holidays <- as.Date(c(
"2021-5-31", "2021-6-19", "2021-7-4",

"2021-9-6", "2021-11-2", "2021-11-25", "2021-11-26", "2021-12-24", "2021-12-25", "2021-12-26", "2021-12-27", "2021-12-28", "2021-12-29", "2021-12-30", "2021-12-31", "2022-1-1", "2022-1-2", "2022-1-17", "2022-2-18", "2022-2-21", "2022-3-18", "2022-4-11", "2022-4-12", "2022-4-13", "2022-4-14", "2022-4-15", "2022-4-16", "2022-4-17", "2022-4-18", "2022-5-30", "2022-6-19", "2022-7-4",

"2022-9-5", "2022-10-5", "2022-10-21", "2022-11-8", "2022-11-9", "2022-11-23", "2022-11-24", "2022-11-25", "2022-12-23", "2022-12-24", "2022-12-25", "2021-12-26", "2022-12-27", "2022-12-28", "2022-12-29", "2022-12-30", "2022-12-31", "2023-01-01", "2023-1-2", "2023-1-16", "2023-1-23", "2023-2-17", "2023-2-20", "2023-3-8", "2023-3-17", "2023-4-3", "2023-4-4", "2023-4-5", "2023-4-6", "2023-4-7", "2023-4-8", "2023-4-9", "2023-4-10", "2023-4-21", "2023-5-29",  "2023-6-19", "2023-7-4", 

"2023-9-4", "2023-10-20", "2023-11-22", "2023-11-23", "2023-11-24", "2023-12-22", "2023-12-23", "2023-12-24", "2023-12-25", "2023-12-26", "2023-12-27", "2023-12-28", "2023-12-29", "2023-12-30", "2023-12-31"))

#For adding the holidays column into the data frame
df3 <- df3 %>%
  mutate(holiday_bin = ifelse(date %in% holidays, 1, 0))

#-------#

#dayOfTheWeek Holiday Results


# split these into 2 for winter non winter months

model2<-felm(call_bin ~ temp_over_100 + temp_95_100 + temp_90_95 + temp_85_90 + temp_80_85 + temp_75_80 + temp_70_75 + temp_65_70 + temp_60_65 + temp_under_60 + holiday_bin + holiday_bin:daytime + daytime + temp_over_100:daytime + temp_95_100: daytime + temp_90_95: daytime + temp_85_90: daytime + temp_80_85: daytime + temp_75_80: daytime + temp_70_75: daytime + temp_65_70: daytime + temp_60_65: daytime + temp_under_60| 
               policeDistrict + year + month + dow, data=df3)

model3<-felm(call_bin ~ temp_over_100 + temp_95_100 + temp_90_95 + temp_85_90 + temp_80_85 + temp_75_80 + temp_70_75 + temp_65_70 + temp_60_65 + temp_under_60 + daytime + holiday_bin + holiday_bin:daytime + temp_over_100:daytime + temp_95_100: daytime + temp_90_95: daytime + temp_85_90: daytime + temp_80_85: daytime + temp_75_80: daytime + temp_70_75: daytime + temp_65_70: daytime + temp_60_65: daytime + temp_under_60| 
               policeDistrict + year + dow, data=df3)

summary(model2)
summary(model3)
```

For the Map:

```{r}
#| echo: true

library("terra")

# baltimore_city<-read.csv("baltimore_zipcodes.csv")

# baltimore <- c("21201", "21202", "21205", "21206", "21207", "21208", "21209", "21210", "21211", "21212", "21213", "21214", "21215", "21216", "21217", "21218", "21222", "21223", "21224", "21225", "21226", "21227", "21228", "21229", "21230", "21231", "21234", "21236", "21237", "21239", "21251", "21287")

# baltimore_city$trt <- 0

# plot(baltimore_city, "trt", col=map.pal("blues"))
```

## Question 10: What are the next steps of your research?

Future Plans:

-   Alter the graph so that it shows the relationship between specific temperature periods to the number of calls

-   Incorporate methods and strategies from official articles into our research paper.

-   Control for trees, precipitation and air quality

-   Identify factors that impacts temperature and number of calls at the same time.

    -   precipitation

    -   trees

    -   air quality

Step 9: Change the document format to gfm

Step 10: Save this document as README.qmd

Step 11: Render the document. README.md file should be created after this process.

Step 12: Push the document back to GitHub and observe your beautiful document in your repository!

Step 13: If your team has a complete dataframe that includes both the treated and outcome variable, you are done with the assignment. If not, make a research plan in Notion to collect data on the outcome and treatment variable and combine it into one dataframe.
