---
title: "How does high tempuratures affect mental health in Baltimore City?"
format: gfm
editor: visual
---

Step 1. Install necessary packages.

```{r}
#| eval: false
install.packages("tidyverse")
install.packages("kableExtra")
```

Step 2. Declare that you will use these packages in this session.

```{r}
library("tidyverse")
library("kableExtra")
```

Step 3. Upload the dataframe that you have created in Spring 2024 into the repository.

Step 4. Open the dataframe into the RStudio Environment.

```{r}
df<-read.csv("panel.csv")
df2<-df %>%
  mutate(call_bin = ifelse(callscount>0, 1, 0)) %>%
  mutate(date = as.Date(date)) %>%
  mutate(month = month(date)) %>%
  mutate(dow = weekdays(date))
```

Step 5. Use the **head** and **kable** function showcase the first 10 rows of the dataframe to the reader.

```{r}
kable(head(df2))
```

## Question 1: What is the frequency of this data frame?

Answer: Daily- day and night

## Question 2: What is the cross-sectional (geographical) unit of this data frame?

Answer: Police district

Step 6. Use the **names** function to display all the variables (column) in the dataframe.

```{r}
names(df2)

```

## Question 3: Which column represents the treatment variable of interest?

Answer: temp_F

## Question 4: Which column represents the outcome variable of interest?

Answer: call_bin

Step 7: Create a boxplot to visualize the distribution of the outcome variable under treatment and no treatment.

```{r}
# New data frame with a column that stores the tempurature ranges.
dfTest <- df2 %>%
  mutate(tempCatagories = cut(temp_F, 
                              breaks = c(-Inf, 60, 65, 70, 75, 80, 85, 90, 95, 100, Inf), 
                              labels = c("< 60", "60-65", "65-70", "70-75", "75-80", 
                                         "80-85", "85-90", "90-95", "95-100", "> 100"),
                              right = TRUE))
#ggplot(df2, aes(x=temp_F)) +
  #geom_boxplot() +
  #facet_wrap(~call_bin)


# Summarize the data by tempCatagories and year
dfTest <- dfTest %>%
  group_by(tempCatagories, year) %>%
  summarise(
    calls = sum(call_bin),          # Total number of calls
    days = n()                       # Count the number of days in each group
  ) %>%
  mutate(avg_calls = calls / days)   # Calculate average calls per day

# Create the line plot with multiple lines for each year
ggplot(dfTest, aes(x = tempCatagories, y = avg_calls, group = year, color = as.factor(year))) +
  geom_line(size = 1) +
  geom_point() + # Optional: Adds points to the lines for better visibility
  labs(title = "Average Number of Calls per Temperature Range (by Year)",
       x = "Temperature Categories",
       y = "Average Number of Calls",
       color = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate the x-axis labels for readability
```

![](data:image/png;base64,iVBORw0KGgoAAAANSUhEUgAAACgAAAAaCAYAAADFTB7LAAAAcElEQVR4Xu3OwQmAQAxE0bClWYCW5N06tM6V2YPg5CjoF/JhLoHAi6iqn9eOefUbqrYvHY0cQDLyAlKRNyARmYA0ZMLRkAlGQyaU72tkAtlim7r/vJqDUDjlKBROOQyFU2icQuMUGqfQuBEaV1XPOwEx96nYACK8+wAAAABJRU5ErkJggg== "Run Current Chunk")

Step 8: Fit a regression model $y=\beta_0 + \beta_1 x + \epsilon$ where $y$ is the outcome variable and $x$ is the treatment variable. Use the **summary** function to display the results.

```{r}
model1<-lm(call_bin ~ temp_F, data=df2)

summary(model1)
```

## Question 7: What is the equation that describes the linear regression above? Please include an explanation of the variables and subscripts.

Answer:

$$
callscount_{pdt} = \beta_0 + \beta_1 temp_{pdt} + \gamma_p + \eta_d + \theta_{dayofweek} + \omega_{month} + \pi_year+\epsilon_{pdt}
$$

Where $callscount$ represents the outcome variable that shows whether there a call from police district \$p\$, on day \$d\$, at time of day \$t\$.

Where $temp$ determines the temperature in Fahrenheit.

## Question 8: What fixed effects can be included in the regression? What does each fixed effects control for? Please include a new equation that incorporates the fixed effects.

Answer:

Police district: controls for better or worse conditions in the city

Daytime: controls the differences that might occur based on the time of day

Year: controls any differences that might have occurred one year to another

## Question 9: What is the impact of the treatment effect once fixed effects are included?

Answer:

```{r}

#install.packages("lfe")
library("lfe")

model2<-felm(call_bin ~ temp_F + daytime + temp_F:daytime| 
               policeDistrict + year + month + dow, data=df2)

summary(model2)
```

```{r}

df3<-df2 %>%
  mutate(temp_over_100=ifelse(temp_F>100, 1, 0)) %>%
  mutate(temp_95_100=ifelse(temp_F> 95 & temp_F<=100, 1, 0)) %>%
  mutate(temp_90_95=ifelse(temp_F> 90 & temp_F<=95, 1, 0)) %>%
  mutate(temp_85_90=ifelse(temp_F> 85 & temp_F<=90, 1, 0)) %>%
  mutate(temp_80_85=ifelse(temp_F> 80 & temp_F<=85, 1, 0)) %>%
  mutate(temp_75_80=ifelse(temp_F> 75 & temp_F<=80, 1, 0)) %>%
  mutate(temp_70_75=ifelse(temp_F> 70 & temp_F<=75, 1, 0)) %>%
  mutate(temp_65_70=ifelse(temp_F> 65 & temp_F<=70, 1, 0)) %>%
  mutate(temp_60_65=ifelse(temp_F> 60 & temp_F<=65, 1, 0)) %>%
  mutate(temp_under_60=ifelse(temp_F<= 60, 1, 0))

model2<-felm(call_bin ~ temp_over_100 + temp_95_100 + temp_90_95 + temp_85_90 + temp_80_85 + temp_75_80 + temp_70_75 + temp_65_70 + temp_60_65 + temp_under_60 +
               daytime + 
               temp_over_100:daytime + temp_95_100: daytime + temp_90_95: daytime + temp_85_90: daytime + temp_80_85: daytime + temp_75_80: daytime + temp_70_75: daytime + temp_65_70: daytime + temp_60_65: daytime + temp_under_60| 
               policeDistrict + year + month + dow, data=df3)

summary(model2)
```

## Question 10: What are the next steps of your research?

-   Identify factors that impacts temperature and number of calls at the same time.

    -   precipitation

    -   trees

    -   air quality

Step 9: Change the document format to gfm

Step 10: Save this document as README.qmd

Step 11: Render the document. README.md file should be created after this process.

Step 12: Push the document back to GitHub and observe your beautiful document in your repository!

Step 13: If your team has a complete dataframe that includes both the treated and outcome variable, you are done with the assignment. If not, make a research plan in Notion to collect data on the outcome and treatment variable and combine it into one dataframe.

Data Description:

-   Outcome variable: whether there was a call or not per day.

-   Treatment variable: temperature

-   Frequency and geographical Unit: daily day and night and police district

-   Treated Group: \# of call on a hot weekday in a specific month in a police district

-   Untreated Group:# of call on a cool weekday in a specific month in a police district

Future Plans:

-   Alter the graph so that it shows the relationship between specific temperature periods to the number of calls

-   Incorporate methods and strategies from official articles into our research paper.

-   Control for trees, precipitation and air quality
